Hydra Battles and Cie

### 2021-06-29 doc/chapter-powers.tex
consistent use of capitalization in headers
fix link in pdf

\chapter{Smart Computation of \texorpdfstring{$x^n$}{Powers}}
\label{chapter-powers}
\section{Introduction}

@@

\emph{From Module 
\href{../theories/html/hydras.additions.FirstSteps.html}{\texttt{additions.FirstSteps}}}
\label{sect: power-definitions}

@@

\emph{From Module
\href{../theories/html/hydras.additions.FirstSteps.html}{additions.FirstSteps}}

@@

\emph{From Library
\href{../theories/html/hydras.additions.Fib2.html}{additions.Fib2}}

@@

\emph{From Module \href{../theories/html/hydras.additions.Monoid_def.html}{additions/Monoid\_def.v}.}

@@

\emph{From Module \href{../theories/html/hydras.additions.Monoid_def.html}{additions.Monoid\_def}}

@@

The following class definition, from Module \href{../theories/html/hydras.additions.Monoid_def.html}{additions.Monoid\_def},
is parameterized with some type $A$,
a multiplication (called \texttt{op} in the definition), and a neutral element
$\mathds{1}$ (called \texttt{one} in the definition).

@@

Let us show various instances, which will be used in further proofs and examples.
Complete definitions and proofs are given in 
File~\href{../theories/html/hydras.additions.Monoid_instances.html}{additions/Monoid\_instances.v}.

@@

\noindent
\emph{From Module \href{../theories/html/hydras.additions.Monoid_def.html}{additions.Monoid\_def}}

@@

\emph{From Module \href{../theories/html/hydras.additions.Monoid_instances.html}{additions.Monoid\_instances}}

@@

The  module \href{../theories/html/hydras.additions.Pow.html}{additions.Pow} defines two functions for exponentiation on any 
\texttt{EMonoid}  on carrier $A$.
They are essentially the same as in Sect.~\vref{sect: power-definitions}. The main difference lies in the arguments of the functions, which now contain
 an instance~\texttt{M} of class \texttt{EMonoid}. 
Thus, the arguments associated with the multiplication,
the neutral element and the equivalence relation associated with \texttt{M}
are left implicit.

@@

From Module~\href{../theories/html/hydras.additions.Demo_power.html}{additions.Demo\_power}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.additions.Pow.html}{additions.Pow}}

@@

\vspace{4pt}
From Module~\href{../theories/html/hydras.additions.Pow.html}{additions.Pow}

@@

\subsection{Proof of correctness of binary exponentiation w.r.t. the function \texttt{power}}
Section \texttt{M\_given} of Module 
~\href{../theories/html/hydras.additions.Pow.html}{additions.Pow} is devoted to the proof 
of properties of the functions above.
Note that properties of \texttt{power} refer to the \emph{specification} of exponentiation, and can be applied for proving correctness of any implementation.

@@

From Module~\href{../theories/html/hydras.additions.Addition_Chains.html}{additions.Addition\_Chains}

@@

\textbf{Note} A first solution (in ~\href{../theories/html/hydras.additions.Trace_exercise.html}{additions.Trace\_exercise}) consists in the definition of 
a (non-associative) multiplication over a type of trace, and apply the function
\texttt{chain\_execute} as if it were computing a power of \texttt{(1,Init)}.

@@

This tactic is not adapted to much bigger exponents. In \linebreak
 Module~\href{../theories/html/hydras.additions.Euclidean_Chains.html}{Euclidean\_Chains},
 for instance, we tried to apply this tactic for proving the correctness 
of a chain associated with the exponent $45319$. 
 We had to interrupt the prover, which 
was trying to build a linear tree of $2\times  45319 + 1$ nodes!
Indeed, using \texttt{reflection\_correct\_tac} is like doing a 
symbolic evaluation of an inefficient (linear) exponentiation algorithm.

@@

\paragraph{Note:}
All the \coq{} material described in this section is available on 
 Module~\href{../theories/html/hydras.additions.Euclidean_Chains.html}{additions/Euclidean\_Chains.v}

@@

\vspace{4pt}
\noindent
From ~\href{../theories/html/hydras.additions.Dichotomy.html}{additions.Dichotomy}.
\begin{Coqsrc}
Class Strategy (gamma : positive -> positive):=
{
gamma_lt :forall p:positive, 3 < p -> gamma  p < p;
gamma_gt : forall p:positive, 3 < p -> 1 < gamma  p
}.

@@

This strategy is defined in Module ~\href{../theories/html/hydras.additions.Dichotomy.html}{additions.Dichotomy}.

@@

\emph{From Module~\href{../theories/html/hydras.additions.BinaryStrat.html}{additions.BinaryStrat}.}

@@

From Module~\href{../theories/html/hydras.additions.AM.html}{additions.AM}

@@

In the library~\href{../theories/html/hydras.additions.AM.html}{additions.AM},
we define a chain generator for this data structure. 
Please note that many proof scripts are copied verbatim from 
\texttt{Euclidean\_Chains} into \texttt{AM}. Removing such redundancies is left as a project.

>>>>

\chapter{Smart computation of \texorpdfstring{$x^n$}{Powers}}
\label{chapter-powers}
\section{Introduction}

@@

\emph{From Module 
\href{../theories/html/additions.FirstSteps.html}{\texttt{additions.FirstSteps}}}
\label{sect: power-definitions}

@@

\emph{From Module
\href{../theories/html/FirstSteps.html}{additions.FirstSteps}}

@@

\emph{From Library
\href{../theories/html/additions.Fib2.html}{additions.Fib2}}

@@

\emph{From Module \href{../theories/html/additions.Monoid_def.html}{additions/Monoid\_def.v}.}

@@

\emph{From Module \href{../theories/html/additions.Monoid_def.html}{additions.Monoid\_def}}

@@

The following class definition, from Module \href{../theories/html/additions.Monoid_def.html}{additions.Monoid\_def},
is parameterized with some type $A$,
a multiplication (called \texttt{op} in the definition), and a neutral element
$\mathds{1}$ (called \texttt{one} in the definition).

@@

Let us show various instances, which will be used in further proofs and examples.
Complete definitions and proofs are given in 
File~\href{../theories/html/additions.Monoid_instances.html}{additions/Monoid\_instances.v}.

@@

\noindent
\emph{From Module \href{../theories/html/additions.Monoid_def.html}{additions.Monoid\_def}}

@@

\emph{From Module \href{../theories/html/additions.Monoid_instances.html}{additions.Monoid\_instances}}

@@

The  module \href{../theories/html/additions.Pow.html}{additions.Pow} defines two functions for exponentiation on any 
\texttt{EMonoid}  on carrier $A$.
They are essentially the same as in Sect.~\vref{sect: power-definitions}. The main difference lies in the arguments of the functions, which now contain
 an instance~\texttt{M} of class \texttt{EMonoid}. 
Thus, the arguments associated with the multiplication,
the neutral element and the equivalence relation associated with \texttt{M}
are left implicit.

@@

From Module~\href{../theories/html/additions.Demo_power.html}{additions.Demo\_power}

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/additions.Pow.html}{additions.Pow}}

@@

\vspace{4pt}
From Module~\href{../theories/html/additions.Pow.html}{additions.Pow}

@@

\subsection{Proof of correctness of binary exponentiation w.r.t. the function \texttt{power}}
Section \texttt{M\_given} of Module 
~\href{../theories/html/additions.Pow.html}{additions.Pow} is devoted to the proof 
of properties of the functions above.
Note that properties of \texttt{power} refer to the \emph{specification} of exponentiation, and can be applied for proving correctness of any implementation.

@@

From Module~\href{../theories/html/additions.Addition_Chains.html}{additions.Addition\_Chains}

@@

\textbf{Note} A first solution (in ~\href{../theories/html/additions.Trace_exercise.html}{additions.Trace\_exercise}) consists in the definition of 
a (non-associative) multiplication over a type of trace, and apply the function
\texttt{chain\_execute} as if it were computing a power of \texttt{(1,Init)}.

@@

This tactic is not adapted to much bigger exponents. In \linebreak
 Module~\href{../theories/html/additions.Euclidean_Chains.html}{Euclidean\_Chains},
 for instance, we tried to apply this tactic for proving the correctness 
of a chain associated with the exponent $45319$. 
 We had to interrupt the prover, which 
was trying to build a linear tree of $2\times  45319 + 1$ nodes!
Indeed, using \texttt{reflection\_correct\_tac} is like doing a 
symbolic evaluation of an inefficient (linear) exponentiation algorithm.

@@

\paragraph{Note:}
All the \coq{} material described in this section is available on 
 Module~\href{../theories/html/additions.Euclidean_Chains.html}{additions/Euclidean\_Chains.v}

@@

\vspace{4pt}
\noindent
From ~\href{../theories/html/additions.Dichotomy.html}{additions.Dichotomy}.
\begin{Coqsrc}
Class Strategy (gamma : positive -> positive):=
{
gamma_lt :forall p:positive, 3 < p -> gamma  p < p;
gamma_gt : forall p:positive, 3 < p -> 1 < gamma  p
}.

@@

This strategy is defined in Module ~\href{../theories/html/additions.Dichotomy.html}{additions.Dichotomy}.

@@

\emph{From Module~\href{../theories/html/additions.BinaryStrat.html}{additions.BinaryStrat}.}

@@

From Module~\href{../theories/html/additions.AM.html}{additions.AM}

@@

In the library~\href{../theories/html/additions.AM.html}{additions.AM},
we define a chain generator for this data structure. 
Please note that many proof scripts are copied verbatim from 
\texttt{Euclidean\_Chains} into \texttt{AM}. Removing such redundancies is left as a project.

### 2021-06-29 doc/part-hydras.tex
fix link in pdf

\vspace{4pt}
\noindent\emph{From Library~\href{../theories/html/hydras.OrdinalNotations.Definitions.html}{OrdinalNotations.Definitions}}

@@

\emph{You may start this exercise with the file
\url{../exercises/ordinals/predSuccUnicity.v}.}

@@

\emph{You may start this exercise with the file
\url{../exercises/ordinals/lt_succ_le.v}.}
\end{exercise}

@@

The detailed  proof script is in the file \url{../theories/html/hydras.Hydra.Omega2_Small.html}.

@@

\emph{You may start this exercise with the file
\url{../exercises/ordinals/Limit_Infinity.v}.}
 \end{exercise}

@@

\emph{You may start this exercise with the file
\url{../exercises/ordinals/ge_omega_iff.v}.}
\end{exercise}

@@

\emph{You may start this exercise with the file
\url{../exercises/ordinals/T1_ltNotWf.v}.}
\end{exercise}

@@

The following lemma is an application of the strict inequality 
\showmath{\alpha < \omega ^\alpha}. If \showmath{\alpha} is strongly accessible, then, by definition, \showmath{\omega^\alpha} is accessible, thus \showmath{\alpha} is \emph{a fortiori} accessible.

@@

Thus, it remains to prove that every ordinal strictly less than \showmath{\epsilon_0} 
is strongly accessible.

@@

First, we prove that, for  any \texttt{LT}-accessible term \showmath{\alpha}, \showmath{\alpha} is 
strongly accessible too (\emph{i.e.} any well formed
term (\texttt{ocons $\alpha$ $n$ $\beta$})  is accessible).

@@

The proof is structured as an induction on \showmath{\alpha}'s accessibility. Let us consider
any  accessible term $\alpha$.

@@

\begin{remark}
\label{remark:a3pat}
The alternate proof of well-foundedness using \'Evelyne Contejean's work on recursive path ordering~\cite{DershowitzRPO, a3pat} is available in the library \href{../theories/html/hydras.Epsilon0.Epsilon0rpo.html}{Epsilon0rpo}.
 \end{remark}

@@

\emph{From Module~\href{../theories/html/hydras.Schutte.Schutte.Correctness_E0.html}{Schutte.Schutte.Correctness\_E0}}

@@

\textbf{Hint}. You may use the following definition (in 
\href{../theories/html/hydras.OrdinalNotations.Definitions.html}{OrdinalNotations.Definitions}).

@@

\end{project}
\section{A Variant for hydra battles}

@@

In the module \href{../theories/html/hydras.Hydra.Epsilon0_Needed_Generic.html}{Epsilon0\_Needed\_Generic}, we assume there exists some variant $m$ bounded by some ordinal $\mu<\epsilon_0$. This part of the development is parameterized by some class $B$ of battles, which will be instantiated later to \texttt{free} or \texttt{standard}.

@@

\vspace{4pt}
\emph{From Module~\href{../theories/html/hydras.Epsilon0.KS.html}{Epsilon0.KS}}

@@

\subsection{Back to hydras}
\label{sec:standard-battles-cases}
We are now able to complete our proof that there exists no bounded variant for proving the termination of standard hydra battles. This proof can
be consulted in the module 
\url{../theories/html/hydras.Hydra.Epsilon0_Needed_Std.html}.
Please note that it has the same global structure as in section\ref{sec:free-battles-case} 
% ICI !
Applying the  lemmas  \texttt{Lemma2\_6\_1} of the module 
\href{../theories/html/hydras.Epsilon0.Paths.html\#Lemma2_6_1}%
{\texttt{Epsilon0.Paths}}   and 
\href{../theories/html/hydras.Epsilon0.Paths.html\#constant_to_standard_path}%
{\texttt{constant\_to\_standard\_path}},
we can convert any inequality $\beta<\alpha<\epsilon_0$ into a standard path from
$\alpha$ to  $\beta$, then into a fragment of a standard battle from 
$\iota(\alpha)$ to $\iota(\beta)$, hence the inequality $m(\iota(\beta))<m(\iota(\alpha))$.

@@

On may ask whether the preceding results are too restrictive, since they 
refer to a particular data type \texttt{T1}.
In fact, our representation of ordinals strictly less than 
 $\epsilon_0$ is faithful to their mathematical definition, at least 
Kurt Schütte's~\cite{schutte}, as proved in Chapter~\vref{chap:schutte}.
(please see also \href{../theories/html/hydras.Schutte.Correctness_E0.html}{the module \texttt{hydras.Schutte.Correctness\_E0}}).

@@

The following code comes from 
 \url{../theories/html/hydras.Epsilon0.F_alpha.html}.

@@

\emph{You may start this exercise with the file
\url{../exercises/ordinals/F_3.v}.}
\end{exercise}

@@

\emph{You may start this exercise with the file
\url{../exercises/ordinals/is_F_monotonous.v}.}
\end{exercise}

@@

\emph{You may start this exercise with the file
\url{../exercises/ordinals/F_omega.v}. Properties of the Ackermann function are studied in 
\url{../theories/ordinals/MoreAck/Ack.v} and
\url{../theories/ordinals/MoreAck/AcknotPR.v}.}
\end{exercise}

@@

Schütte's first axiom tells that \texttt{lt} is a well order on the set 
\texttt{ordinal} (The  class \texttt{WO} is defined in
Module~\href{../theories/html/hydras.Schutte.Well_Orders.html}%
{\texttt{Well\_Orders.v}}).

@@

We prove the inequality  $\alpha<\omega^\alpha$ whenever $\alpha < \epsilon_0$.
\emph{Note that this condition was implicit in Module \href{../theories/html/hydras.Epsilon0/T1.html\#lt_phi0}%
{\texttt{Epsilon0.T1}}.}

@@

\emph{You may start this exercise with the file
\url{../exercises/ordinals/schutte_cnf_counter_example.v}.}
\end{exercise}

@@

\begin{remark}
The sub-directory \href{../theories/html/hydras.Gamma0.html}%
{\texttt{theories/Gamma0}} contains an (incomplete, still undocumented) implementation of the set of ordinals below $\Gamma_0$, represented in Veblen normal form. 
\end{remark}

@@

Our library 
\href{../theories/html/hydras.Schutte.correctness_E0.html}%
{\texttt{Schutte.Correctness\_E0}} establishes the link between two very different modelizations of ordinal numbers. In other words, it ``validates'' a data structure in terms of
a classical mathematical discourse considered as a model. 
First, we define a function from \texttt{T1} into  \texttt{Ord} by structural recursion.

>>>>

\vspace{4pt}
\noindent\emph{From Library~\href{../theories/html/hydras.OrdinalNotations.ON_mult.html}{OrdinalNotations.ON\_Generic}}

@@

\emph{You may start this exercise with the file
\href{https://github.com/coq-community/hydra-battles/blob/master/exercises/ordinals/predSuccUnicity.v}{exercises/ordinals/predSuccUnicity.v}.}

@@

\emph{You may start this exercise with the file
\href{https://github.com/coq-community/hydra-battles/blob/master/exercises/ordinals/lt_succ_le.v}{exercises/ordinals/lt\_succ\_le.v}.}
\end{exercise}

@@

The detailed  proof script is in the file
\href{https://github.com/coq-community/hydra-battles/blob/master/theories/ordinals/Hydra/Omega2_Small.v}{theories/ordinals/Hydra/Omega2\_Small.v}.

@@

\emph{You may start this exercise with the file
     \href{https://github.com/coq-community/hydra-battles/tree/master/exercises/ordinals/Limit_Infinity.v}{exercises/ordinals/Limit\_Infinity.v}.}
 \end{exercise}

@@

\emph{You may start this exercise with the file
    \href{https://github.com/coq-community/hydra-battles/tree/master/exercises/ordinals/ge_omega_iff.v}{exercises/ordinals/ge\_omega\_iff.v}.}
\end{exercise}

@@

\emph{You may start this exercise with the file
    \href{https://github.com/coq-community/hydra-battles/tree/master/exercises/ordinals/T1_ltNotWf.v}{exercises/ordinals/T1\_ltNotWf.v}.}
\end{exercise}

@@

The following lemma is an application of the strict inequality 
$\alpha < \omega ^\alpha.$ If \alpha is strongly accessible, then, by definition,
$\omega^\alpha$ is accessible, thus \alpha is \emph{a fortiori} accessible.

@@

Thus, it remains to prove that every ordinal strictly less than $\epsilon_0$
is strongly accessible.

@@

First, we prove that, for  any \texttt{LT}-accessible term $\alpha$, $\alpha$ is 
strongly accessible too (\emph{i.e.} any well formed
term (\texttt{ocons $\alpha$ $n$ $\beta$})  is accessible).

@@

The proof is structured as an induction on $\alpha'$s accessibility. Let us consider
any  accessible term $\alpha$.

@@

\begin{remark}
\label{remark:a3pat}
The alternate proof of well-foundedness using \'Evelyne Contejean's work on
    recursive path ordering~\cite{DershowitzRPO, a3pat} is available in the
    library \href{../theories/html/hydras.Epsilon0.Epsilon0rpo.html}{Epsilon0.Epsilon0rpo}.
 \end{remark}

@@

\emph{From Module~\href{../theories/html/hydras.Schutte.Correctness_E0.html}{Schutte.Correctness\_E0}}

@@

\textbf{Hint}. You may use the following definition (in 
    \href{../theories/html/hydras.OrdinalNotations.ON_Generic.html}{OrdinalNotations.ON\_Generic}).

@@

    \end{project}
    \section{A variant for hydra battles}

@@

In the module
\href{../theories/html/hydras.Hydra.Epsilon0_Needed_Generic.html}{Hydra.Epsilon0\_Needed\_Generic}, we assume there exists some variant $m$ bounded by some ordinal $\mu<\epsilon_0$. This part of the development is parameterized by some class $B$ of battles, which will be instantiated later to \texttt{free} or \texttt{standard}.

@@

@@

\subsection{Back to hydras}
\label{sec:standard-battles-cases}
We are now able to complete our proof that there exists no bounded variant for proving the termination of standard hydra battles. This proof can
be consulted in the module 
\href{../theories/html/hydras.Hydra.Epsilon0_Needed_Std.html}{Hydra.Epsilon0\_Needed\_Std}.
Please note that it has the same global structure as in section\ref{sec:free-battles-case} 
% ICI !
Applying the  lemmas  \texttt{Lemma2\_6\_1} of the module 
\href{../theories/html/hydras.Epsilon0.Paths.html\#Lemma2_6_1}%
{Epsilon0.pathS}   and 
\href{../theories/html/hydras.Epsilon0.Paths.html\#constant_to_standard_path}%
{\texttt{constant\_to\_standard\_path}},
we can convert any inequality $\beta<\alpha<\epsilon_0$ into a standard path from
$\alpha$ to  $\beta$, then into a fragment of a standard battle from 
$\iota(\alpha)$ to $\iota(\beta)$, hence the inequality $m(\iota(\beta))<m(\iota(\alpha))$.

@@

On may ask whether the preceding results are too restrictive, since they 
refer to a particular data type \texttt{T1}.
In fact, our representation of ordinals strictly less than 
 $\epsilon_0$ is faithful to their mathematical definition, at least 
Kurt Schütte's~\cite{schutte}, as proved in Chapter~\vref{chap:schutte}.
(please see also the module
\href{../theories/html/hydras.Schutte.Correctness_E0.html}{hydras.Schutte.Correctness\_E0}).

@@

The following code comes from 
 \href{../theories/html/hydras.Epsilon0.F_alpha.html}{Epsilon0.F\_alpha}.

@@

\emph{You may start this exercise with the file
\href{https://github.com/coq-community/hydra-battles/tree/master/exercises/ordinals/F_3.v}{exercises/ordinals/F\_3.v}.}
\end{exercise}

@@

\emph{You may start this exercise with the file
    \href{https://github.com/coq-community/hydra-battles/tree/master/exercises/ordinals/is_F_monotonous.v}{exercises/ordinals/is\_F\_monotonous.v}.}
\end{exercise}

@@

\emph{You may start this exercise with the file
    \href{https://github.com/coq-community/hydra-battles/tree/master/exercises/ordinals/F_omega.v}{exercises/ordinals/F\_Omega.v}.
Properties of the Ackermann function are studied in
    \href{https://github.com/coq-community/hydra-battles/tree/master/theories/ordinals/MoreAck/Ack.v}{theories/ordinals/MoreAck/Ack.v} and
    \href{https://github.com/coq-community/hydra-battles/tree/master/theories/ordinals/MoreAck/AckNotPR.v}{theories/ordinals/MoreAck/AckNotPR.v}
.}
\end{exercise}

@@

Schütte's first axiom tells that \texttt{lt} is a well order on the set 
\texttt{ordinal} (The  class \texttt{WO} is defined in
Module~\href{../theories/html/hydras.Schutte.Well_Orders.html}{Schutte.Well\_Orders.v}).

@@

We prove the inequality  $\alpha<\omega^\alpha$ whenever $\alpha < \epsilon_0$.
\emph{Note that this condition was implicit in Module
\href{../theories/html/hydras.Epsilon0.T1.html\#lt_phi0}{Epsilon0.T1}.}

@@

\emph{You may start this exercise with the file
    \href{https://github.com/coq-community/hydra-battles/tree/master/exercises/ordinals/schutte_cnf_counter_example.v}{exercises/ordinals/schutte\_cnf\_counter\_example.v}.}
\end{exercise}

@@

\begin{remark}
The sub-directory
    \href{https://github.com/coq-community/hydra-battles/tree/master/theories/ordinals/Gamma0}{theories/oridinals/Gamma0} contains an (incomplete, still undocumented) implementation of the set of ordinals below $\Gamma_0$, represented in Veblen normal form. 
\end{remark}

@@

Our library 
\href{../theories/html/hydras.Schutte.Correctness_E0.html}%
{\texttt{Schutte.Correctness\_E0}} establishes the link between two very different modelizations of ordinal numbers. In other words, it ``validates'' a data structure in terms of
a classical mathematical discourse considered as a model. 
First, we define a function from \texttt{T1} into  \texttt{Ord} by structural recursion.
